---
title: "HW2"
author: "Chenxi Liu 1010615050"
date: "2024-01-19"
output: pdf_document
---

## Lab Exercise 2


```{r}
library(opendatatoronto)
library(tidyverse)
library(stringr)
library(skimr) # EDA
library(visdat) # EDA
library(janitor)
library(lubridate)
library(ggrepel)
```

```{r}
res <- list_package_resources("996cfe8d-fb35-40ce-b569-698d51fc683b") # obtained code from searching data frame above
res <- res |> mutate(year = str_extract(name, "202.?"))
delay_2022_ids <- res |> filter(year==2022) |> select(id) |> pull()

delay_2022 <- get_resource(delay_2022_ids)
# make the column names nicer to work with
delay_2022 <- clean_names(delay_2022)

delay_codes <- get_resource("3900e649-f31e-4b79-9f20-4731bbfd94f7")
delay_data_codebook <- get_resource("ca43ac3d-3940-4315-889b-a9375e7b8aa4")

head(delay_2022)
```

## Question1

```{r}
delay_2022 <- delay_2022 |> distinct()

## Removing the observations that have non-standardized lines
delay_2022 <- delay_2022 |> filter(line %in% c("BD", "YU", "SHP", "SRT"))

delay_2022 %>% 
  filter(min_delay>0) %>%
  group_by(line,station) %>%
  summarise(mean_delay = mean(min_delay)) %>%
  arrange(-mean_delay) %>%
  slice(1:5) %>%
  ggplot(aes(x = station, y = mean_delay)) +
           geom_col() +
           facet_wrap(vars(line), 
             scales = "free_y",
             nrow = 4) +
          coord_flip()
  
```


## Question2

```{r}
delay_2022 <- delay_2022 |> 
  left_join(delay_codes |> rename(code = `SUB RMENU CODE`, code_desc = `CODE DESCRIPTION...3`) |> select(code, code_desc)) 


delay_2022 <- delay_2022 |>
  mutate(code_srt = ifelse(line=="SRT", code, "NA")) |> 
  left_join(delay_codes |> rename(code_srt = `SRT RMENU CODE`, code_desc_srt = `CODE DESCRIPTION...7`) |> select(code_srt, code_desc_srt))  |> 
  mutate(code = ifelse(code_srt=="NA", code, code_srt),
         code_desc = ifelse(is.na(code_desc_srt), code_desc, code_desc_srt)) |> 
  select(-code_srt, -code_desc_srt)

# Calculate the most frequent delay reasons
top_delay_reasons <- delay_2022 %>%
  filter((min_delay > 0) & !is.na(code_desc)) %>%
  count(code_desc, sort = TRUE) %>%
  filter(row_number() <= n() / 2) %>%
  pull(code_desc)
 
# Filter delay_2022 based on conditions
filtered_delay_2022 <- delay_2022 %>%
  filter(min_delay > 0, code_desc %in% top_delay_reasons)

library(MASS)

#Apply the negative binomial regression model
model_nb <- glm.nb(min_delay ~ line + code_desc, data = filtered_delay_2022)

# Summarize the model
summary(model_nb)

```

### Answer2

The results of the Negative Binomial Regression model show that both line and reasons have significant effects on the count of delay minutes.For example, such as "ATC Operator Related" and "ATC Project," are associated with a substantial increase in delay minutes, while delays related to "Assault / Patron Involved" are associated with a decrease in delay minutes. There are also some covariates that do not have significant effects on the delay time, such as "Radio System Fault" and "Track Switch Failure - Signal Related Problem ", which also in line with the reality .


## Question3

```{r}
res1 <- list_package_resources("e869d365-2c15-4893-ad2a-744ca867be3b") # obtained code from searching data frame above
res1 <- res1 |> mutate(year = str_extract(name, "201.?"))
campaign2014_ids <- res1 |> filter(year==2014 & grepl("Data", name)) |> pull(id)
campaign_2014 <- get_resource(campaign2014_ids)

mayor_data <- campaign_2014[grep("Mayor", names(campaign_2014))][[1]]#select the maylor selection related data
colnames(mayor_data) <- unlist(mayor_data[1, ]) #first row as the column names
mayor_data <- mayor_data[-1, ]

mayor_data <- mayor_data %>%
  clean_names()  
head(mayor_data)
```

## Question4

```{r}
skim(mayor_data)

#Check the missing value
mayor_data %>%
  summarize(across(everything(), ~ sum(is.na(.x))))

#Check the duplicates
get_dupes(mayor_data)
mayor_data <- mayor_data |> distinct()

#Check the data type of every column
sapply(mayor_data, class)

#Transfer the character column format to numeric
mayor_data$contribution_amount <- as.numeric(mayor_data$contribution_amount)
```

### Answer4

There are some empty columns :contributors_address, goods_or_service_desc, relationship_to_candidate, president_business_manager, authorized_representative and ward, so we can exclude these empty columns in our following analysis. On top of these empty columns, there is no missing values for other columns.

## Question5

```{r}
#check the distribution of contribution amount
ggplot(data = mayor_data) + 
  geom_histogram(aes(x = contribution_amount)) + scale_x_log10()

#check if there is outlier of thr contribution amount
ggplot(mayor_data, aes(y = contribution_amount)) +
  geom_boxplot()
#check the outlier
mayor_data %>% 
  arrange(-contribution_amount) %>%
  slice(1:10)

#exclude outlier
mayor_data<- mayor_data %>% filter(contribution_amount <= 10000)
  
#check the distribution of contribution type
ggplot(data = mayor_data) + 
  geom_histogram(aes(x = contribution_amount, fill= contribution_type_desc)) + scale_x_log10()

#check the distribution of contributor type
ggplot(data = mayor_data) + 
  geom_histogram(aes(x = contribution_amount, fill= contributor_type_desc)) + scale_x_log10()

#Show the distribution of contribution amount for different candidates
mayor_data |> 
  group_by(candidate) |> 
  summarise(total_amount = sum(contribution_amount)) |> 
  ggplot(aes(x=reorder(candidate, total_amount), y=total_amount)) + 
  geom_col() + 
  coord_flip()
```

### Answer5

By checking the outliers of the contribution amount, we could see that there are 6 rows of amount which are beyond 10000 dollar, and these contributors are also the candidates themselves, they are : Ford Rob, Ford Doug and Goldkind Ari. And 4 out of 6 outlier contributions are from Ford Rob.

## Question6

```{r}

candidate_data <- mayor_data |> 
  group_by(candidate) |> 
  summarise(total_amount = sum(contribution_amount), mean_amount = mean(contribution_amount), number_contribution = n() ) 

candidate_data %>% 
  arrange(-total_amount) %>%
  slice(1:5) %>%
  ggplot(aes(x= candidate, y=total_amount)) + 
  geom_col() 

candidate_data %>% 
  arrange(-mean_amount) %>%
  slice(1:5) %>%
  ggplot(aes(x= candidate, y=mean_amount)) + 
  geom_col() 

candidate_data %>% 
  arrange(-number_contribution) %>%
  slice(1:5) %>%
  ggplot(aes(x= candidate, y=number_contribution)) + 
  geom_col() 

```

## Question7

```{r}

candidate_data_without_self <- mayor_data %>%
  filter(!grepl(candidate, contributors_name)) %>%
  group_by(candidate) %>%
  summarise(
    total_amount = sum(contribution_amount),
    mean_amount = mean(contribution_amount),
    number_contribution = n()
  )

candidate_data_without_self %>% 
  arrange(-total_amount) %>%
  slice(1:5) %>%
  ggplot(aes(x= candidate, y=total_amount)) + 
  geom_col() 

candidate_data_without_self %>% 
  arrange(-mean_amount) %>%
  slice(1:5) %>%
  ggplot(aes(x= candidate, y=mean_amount)) + 
  geom_col() 

candidate_data_without_self %>% 
  arrange(-number_contribution) %>%
  slice(1:5) %>%
  ggplot(aes(x= candidate, y=number_contribution)) + 
  geom_col() 

```

## Question8

```{r}
# Count the number of contributors who gave money to more than one candidate
contributors_multiple_candidates <- mayor_data %>%
  group_by(contributors_name) %>%
  summarise(num_candidates = n_distinct(candidate)) %>%
  filter(num_candidates > 1) %>%
  count()

contributors_multiple_candidates
```

### Answer8

There are 184 contributors who gave money to more than one candidate.
